European or American? A re-examination of the transatlantic influence over the IMF

### 2025-07-30 manuscript/manuscript.qmd
shortened EU lit review part

Compared to the extensive literature on the United States, scholarly analyses of European influence within the IMF remain relatively limited and fragmented. A number of scholars (such as @smaghi2004; @thacker1999; @miescu2014) emphasize that European representation within the Fund is institutionally dispersed, resulting in a fragmented voice that limits collective influence. In contrast, the United States retains a centralized and coherent presence, thereby maintaining a dominant position in key decision-making processes. Nevertheless, other contributions (such as @keating2024; @lutz2014; @copelovitch2010a; @stone2004) argue that European states possess significant structural and institutional advantages—including leadership appointments, voting weight, agenda-setting capacity, transnational financial exposure, and diplomatic leverage—which enable them to exert substantive influence over IMF lending practices.

First, European countries benefit from long-standing institutional privileges within the IMF’s governance architecture. As noted in the introduction, the Fund’s leadership has traditionally been allocated to Europe under an informal transatlantic arrangement that dates back to the Bretton Woods settlement [@kahler2001]. Collectively, EU member states control over 30% of the IMF’s voting rights, with the United Kingdom, France, and Germany each occupying individual executive board seats [@leech2005a]. Quantitative studies suggest that, under a scenario of unified representation, the EU could potentially attain voting power equivalent to a de facto veto similar to that of the United States [@brandner2009a]. These structural features provide European countries with disproportionate influence relative to their global economic weight.

Second, European countries have played a leading role in shaping policy design and operational implementation within the IMF. @lutz2014 demonstrates that the EU and the European Central Bank were instrumental in embedding austerity-oriented conditionalities—reminiscent of the Washington Consensus—in IMF lending to Central and Eastern European states such as Latvia and Romania. This policy orientation was motivated by the significant exposure of French and German banks to these economies. At the request of EU institutions, IMF conditionality in these cases expanded beyond standard macroeconomic stabilization to encompass fiscal rules, structural adjustment, and financial sector supervision [@hodson2014]. Coordination platforms such as EURIMF (European Representatives to the IMF) and SCIMF (Sub-Committee on the IMF) have enabled EU member states to form cohesive negotiating positions and act collectively to advance shared preferences within the Fund [@lutz2014].

In this context, @copelovitch2010 argues that IMF lending decisions are fundamentally political, with European influence manifesting not only through formal mechanisms such as quota shares, but also through informal channels including diplomatic lobbying, technical staffing appointments, and the selection of mission team leaders. During the Eurozone sovereign debt crisis, major European powers, particularly France and Germany, actively solicited IMF participation in stabilization packages. The resulting “troika” mechanism—comprising the IMF, the European Commission, and the European Central Bank—was emblematic of Europe's dual status as both lender and borrower in the crisis [@veron2016]. However, this arrangement attracted criticism for compromising IMF independence. Some scholars contend that excessive deference to European preferences undermined the Fund’s credibility, and that the reluctance of euro area governments to implement IMF-prescribed reforms was a key driver of the crisis’s deepening [@veron2016].

Third, European influence over IMF lending has often served to advance strategic financial and geopolitical interests. This was particularly evident during the Eurozone crisis, as mentioned earlier, when major European economies—most notably Germany and France—sought to mitigate systemic risks arising from their banking sector’s exposure to highly indebted member states. Fiscal surveillance mechanisms, such as the EU’s 3% deficit ceiling, were complemented by IMF-backed assistance programs aimed at restoring financial stability [@heimberger2025, p. 9]. The initiation of the IMF’s financial assistance to Greece, for example, was undertaken at Germany’s explicit request [@kickert2019a]. Beyond economic stabilization, European policymakers also leveraged IMF involvement to enhance the legitimacy of domestic austerity programs, using the Fund’s reputation as an external technocratic authority to depoliticize painful fiscal adjustments [@kickert2019a].

Moreover, European states have extended their influence to extra-European contexts, particularly in Africa. Countries such as the United Kingdom and France have supported IMF concessional lending to former colonies, often motivated by a combination of commercial and strategic interests, including access to natural resources and the preservation of post-colonial diplomatic ties [@stone2004]. However, this form of influence has not been without controversy. Critics such as @moyo2024 and @cain2022 argue that European-backed IMF programs have promoted neoliberal policy frameworks—including market liberalization and privatization—without adequate consideration of local development priorities, undermining African countries' economic sovereighty and state capacity.

As noted earlier, numerous scholars have questioned both the extent and effectiveness of Europe’s influence within the IMF. First, Europe faces persistent challenges in internal coordination. EU member states are represented through multiple, uncoordinated constituencies within the Fund, resulting in a fragmented and often inconsistent policy voice [@smaghi2004]. This institutional fragmentation is compounded by substantive policy divergences among key member states—for instance, France tends to prioritize pro-growth strategies, while Germany places greater emphasis on fiscal discipline [@miescu2014].

Second, despite Europe’s substantial voting power, the United States continues to enjoy structural advantages, most notably its effective veto power over major IMF decisions and its ability to shape key institutional outcomes [@thacker1999]. Moreover, European influence is often constrained by the need to coordinate positions with other members of the G5, which may necessitate concessions that dilute European preferences or undermine coherence [@copelovitch2010a].

Third, concerns have been raised regarding the democratic legitimacy and global representativeness of European countries within the IMF governance structure. @woodward2024 highlights the disproportionate relationship between the population size of EU member states and their aggregate voting power in the Fund, arguing that this imbalance structurally disadvantages developing countries. @leech2005a further suggests that while a consolidation of European seats could potentially establish a bipolar governance configuration capable of counterbalancing U.S. dominance, such a shift might also open space for increased participation and influence by small and medium-sized states within the IMF’s decision-making processes.

However, while these studies offer important insights, they tend to remain case-specific or anecdotal. Crucially, they stop short of offering a systematic comparison between American and European influence, leaving open the question of how these two centres of power operate within the IMF.

>>>>

Compared to the extensive literature on U.S. dominance in the IMF, scholarship on European influence is relatively limited and often fragmented. Many studies underscore that EU member states are institutionally dispersed across multiple constituencies, hindering coherent collective action [@smaghi2004; @miescu2014; @thacker1999]. Still, other works argue that European countries wield significant structural advantages—including leadership appointments, voting power, and coordination mechanisms—that enable them to shape lending outcomes in line with their interests [@copelovitch2010a; @keating2024; @lutz2014; @stone2004].

European influence reflects a strategic combination of formal power and coordinated action. EU countries collectively control over 30% of IMF voting rights, and if they were to unify, they could potentially counterbalance U.S. veto power [@brandner2009a; @leech2005a; @smaghi2004]. This is reinforced by the long-standing transatlantic norm allocating IMF leadership to a European national [@kahler2001]. Through platforms such as EURIMF and SCIMF, EU members have harmonized policy positions and influenced loan design. During the Eurozone crisis, France and Germany pushed for IMF involvement to stabilize the region while embedding austerity-oriented conditionality that protected their banking systems [@hodson2014; @kickert2019a; @veron2016]. Lending terms were shaped to reflect creditor priorities, with evidence that conditionality was both intensified or relaxed based on geopolitical alignment [@lutz2014; @heimberger2025, p. 9; @dreher2015]. Beyond Europe, the UK and France have used IMF concessional lending in Africa to maintain post-colonial ties and commercial access [@stone2004].

Despite Europe’s institutional advantages, scholars remain skeptical of its actual effectiveness in the IMF. Internal divergences—such as France’s growth-oriented stance versus Germany’s fiscal conservatism—combined with fragmented constituencies, often impede coherent action and reduce policy influence [@miescu2014; @dreher2015; @copelovitch2010a]. Moreover, Europe’s disproportionate voting power relative to its population raises legitimacy concerns and is seen as disadvantaging developing countries [@woodward2024; @leech2005a]. While proposals to consolidate EU seats aim to improve efficiency and counterbalance U.S. dominance, critics warn such reforms may merely reinforce dominance by a few large European states rather than broaden representation [@leech2005a; @copelovitch2010a].

While several studies examine European involvement in IMF operations, most focus on individual cases or regional crises, lacking generalizable conclusions. Existing literature often treats Europe as a secondary actor or subsumes it under undifferentiated “Western” influence, thereby overlooking intra-Western variation. By analytically separating U.S. and European roles, our study addresses these gaps and offers a comparative framework for understanding divergent patterns of influence within global financial governance.
